Supplement: Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2

### 2025-04-15 response.Rmd

---
title: "Response to Reviewers"
author:
  - name: Greg Gloor 
    email: ggloor@uwo.ca 
cite-method: citeproc 
citeproc: TRUE 
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib 
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
---

Dear Editor and reviewers

Thank you all for thoughtful and helpful comments. In response to your queries we have added a new Figure 1 that more clearly articulates (we hope) the need for scale uncertainty. Secondly, we  added in new analyses using semi-synthetic datasets to add in a standard of truth. We did this using some of the datasets of, but in a way that was independent of Li et al [@Li:2022aa] and observed essentially the same result. In doing this analysis it became clear that the addition of scale uncertainty was much better at controlling FP than  fold-change cutoffs and we came to understand why this is so and have included this in the revision.

We have also tried to clean up language and clarify sources of data, etc as requested.

We have included line numbers in many of the responses and have included line numbers in our resubmission for cross-reference.

Finally, we have included a file with the differences (diff.pdf) marked up in the package so that the changes can be identified.

We believe  that this revised document is now much stronger because of the reviews and hope it meets approval

The repository for this version is at: https://github.com/ggloor/scale-sim-bio/ and is now public

Best regards

Greg Gloor, on behalf of the authors

### 2025-04-15 response.tex

\title{Response to Reviewers}
\author{true}
\date{}

\begin{document}
\maketitle

Dear Editor and reviewers

Thank you all for thoughtful and helpful comments. In response to your
queries we have added a new Figure 1 that more clearly articulates (we
hope) the need for scale uncertainty. Secondly, we added in new analyses
using semi-synthetic datasets to add in a standard of truth. We did this
using some of the datasets of, but in a way that was independent of Li
et al (1) and observed essentially the same result. In doing this
analysis it became clear that the addition of scale uncertainty was much
better at controlling FP than fold-change cutoffs and we came to
understand why this is so and have included this in the revision.

We have also tried to clean up language and clarify sources of data, etc
as requested.

We have included line numbers in many of the responses and have included
line numbers in our resubmission for cross-reference.

Finally, we have included a file with the differences (diff.pdf) marked
up in the package so that the changes can be identified.

We believe that this revised document is now much stronger because of
the reviews and hope it meets approval

The repository for this version is at:
\url{https://github.com/ggloor/scale-sim-bio/} and is now public

Best regards

Greg Gloor, on behalf of the authors

@@

Comments to the Author

Gloor, Nixon, and Silverman provide a clear and concisely written
manuscript that describes a significant improvement of the ALDEx2
package to incorporate scale uncertainty in analyses of differential
expression and abundance. These scale models relieve the reliance on
commonly used but difficult to rationalize cutoffs of q-values and
log-fold change thresholds, and facilitate adoption through the ALDEx2 R
package. This work represents a valuable addition to an already
impactful and versatile tool for the community. I have several minor
comments that I am confident can be easily addressed by the authors.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The authors focus here on the application of scale-informed ALDEx2 to
  transcriptomic analysis. Can the authors further discuss whether
  applications to other types of measurements is equally appropriate? If
  so, there may be an opportunity to generalize the title to use
  ``counts'' instead of ``RNA-sequencing'', as that broadens
  applicability to other data types (e.g.~proteomics, single cell
  epigenome data, etc).

  \begin{itemize}
  \tightlist
  \item
    We have changed the title to reflect that it is count data, and have
    included information on other data types in the introduction.
    However, the focus of this report is intended to be squarely on
    transcriptome and metatranscriptome data in an attempt to more
    tightly target that analytic audience. Companion articles by by
    McGovern et al. (2) discussing the use in gene set enrichment and by
    Nixon et al. (3) targeting the microbiome community are published in
    PLoS Comp. Bio. and under second revision at Genome Biol.
    respectively.
  \end{itemize}
\item
  Relevant to statements like this: ``The original naive ALDEx2 (39)
  model unwittingly made a strict assumption about scale through the CLR
  normalization (3).'' - Can ``scale-naïve'' be used instead to more
  precisely indicate the distinction between the previous and current
  defaults in ALDEx2?

@@

Comments to the Author Summary: In this manuscript Gloor et
al.~highlight the application of ALDEx2 and the recent addition of scale
models to help improve differential expression analysis in RNA seq
experiments. This manuscript builds upon previous work that introduced
scale models to ALDEx2 and highlights that differential analysis in high
throughput sequencing often suffers from scale assumptions introduced
during commonly applied normalization techniques. In this manuscript
they used both a transcriptomic and metatranscriptomic experiment to
highlight the improvements that can be achieved by using the scale
models incorporated in ALDEx2. The manuscript is well written and
highlights an important problem within the field, however, there are
some major comments that should be addressed by the authors.

Major comments:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Throughout the manuscript a clear definition of what the ``truth'' is
  for each experimental set up is needed (i.e., is it transcripts
  relative to species abundance or bulk absolute transcripts or
  something else?). This is critically important in the
  metatranscriptomics section where high levels of differing biomass
  could result in higher levels of absolute transcripts across the board
  but not higher levels of transcripts relative to the underlying
  species abundance. The attempt to center the data on housekeeping
  genes suggests the latter is the objective but this is not clear
  throughout the manuscript. In addition to this if the authors are
  interested in identifying the transcript levels relative to species
  abundance, they may be interested in comparing their tests against a
  method that normalizes transcript levels against underlying DNA
  abundances(\url{https://pubmed.ncbi.nlm.nih.gov/34465175/}).

  \begin{itemize}
  \tightlist
  \item
    We have added two semi-synthetic datasets to show how the FDR is
    controlled much better by scale uncertainty than by fold-change
    cutoffs. Ultimately, the choice of what to use as the baseline for
    normalization of scale between samples is a choice that must be made
    by the analyst. In some situations the biological question may be
    best addressed by normalizing transcript levels to species abundance
    (although to be clear, in the method suggested both the RNA and
    species levels are relative abundances). However, such a
    normalization is discussed with reference to species-level
    metatranscriptomics rather than systems-level metatranscriptomics as
    shown in our report and in (4). We would need to redo all of the
    analysis in the cited paper in order to make this comparison.
    Indeed, the scale approach advocated in this report can use any
    measure of absolute abundance as an anchor and we have been explicit
    throughout about this. (lines 177-237, 377-380)
  \end{itemize}
\item
  The claims in Figure 1 should be further validated through simulated
  data and or a dataset where the true underlying number of
  differentially expressed transcripts is known. While the reduction in
  transcripts that the scale model achieves is likely inline with the
  biology it is unclear if the model is being overconservative.

  \begin{itemize}
  \tightlist
  \item
    We thank the reviewer for this suggestion while the original paper
    by Nixon et al. (5) included extensive simulation and theoretic
    justification we recognize that this is not yet published and a
    clear demonstration of the utility in transcriptome datasets would
    be helpful. We have now included two simulated datasets from a
    recent benchmarking paper {[}2Li:2022aa{]} showing that the
    scale-reliant approach is substantially better in controlling the
    FDR and, at least in one example, has minimal effect on power (Fig
    2, Supp Fig 3). This has resulted in a major re-write of the results
    section and has made a substantial improvement for how we understand
    scale uncertainty and its effect on data analys. We again thank the
    reviewer for pushing us to do this analysis. (lines 177-237)
  \end{itemize}
\item
  In the metatranscriptomics section it is claimed that biomass differs
  by 20 fold between the groups that are being compared, however, the
  scaled model only includes a difference of 14\%. Please address this
  discrepancy. Furthermore, based on Figure 3B, the gamma model by
  itself was ineffective in this setting, so the manuscript should
  clarify when only the gamma parameter versus the gamma and mu
  parameters must be informed.

  \begin{itemize}
  \tightlist
  \item
    We have tried to include more insights in the results (Supp Figure
    4) and the discussion around when to use an informed vs a simple
    scale model. In essence, the centre of mass of the difference
    between groups should be close to 0 (no difference) to avoid FP and
    FN (6). (lines 358-365, 398-410)
  \end{itemize}
\item
  Due to the limited number of datasets shown in this manuscript the
  robustness of the model for various biological and technical
  conditions is difficult to assess.
\end{enumerate}

@@

  \begin{itemize}
  \tightlist
  \item
    Nixon et al.~show that in theory, any interpretable log-ratio will
    work with this schema and other papers are in preparation showing
    this. For the purposes of this manuscript the CLR is the
    normalization used by ALDEx2
  \end{itemize}
\end{enumerate}

Associate Editor: Erb, Ionas Comments to the Author: Dear authors,

Please focus your revision on the following aspects: State what the true
scale of the data should describe, i.e., number of transcripts relative
to species abundance or similar. Does this change between data sets?
Include a comparison with other approaches, e.g., comparing with an
approach that uses DNA abundance as described in the paper in Annual
Reviews (see reviewer 2).

\begin{itemize}
\tightlist
\item
  we have added a paragraph in the discussion about this. Fundamentally,
  while the scale approach is generalizable and could be applied to the
  types of data given in the review paper (the actual method is in (7)),
  but that approach is designed to ask a different type of question than
  is addressed by scale. (Lines 374-380 and above)
\end{itemize}

@@

Discuss effect of biological replicates vs.~technical replicates (see
both reviewers).

\begin{itemize}
\tightlist
\item
  This has been added in multiple places and again the introduction of
  the simulated dataset has helped with this point.
\end{itemize}

Please indicate if the preprints referenced in (3) and (20) are
submitted / currently under review.

\begin{itemize}
\tightlist
\item
  The companion articles by by McGovern et al. (2) discussing the use in
  gene set enrichment and by Nixon et al. (3) targeting the microbiome
  community are published in PLoS Comp. Bio. and under second revision
  at Genome Biol. respectively. The companion article introducing the
  theory and the initial modeling by Nixon et al. (5) is still under
  review at Annals of applied Statistics which is a frustration for the
  authors as much as the reviewers
\end{itemize}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-Li:2022aa}
1. Li,Y., Ge,X., Peng,F., Li,W. and Li,J.J. (2022)
\href{https://doi.org/10.1186/s13059-022-02648-4}{Exaggerated false
positives by popular differential expression methods when analyzing
human population samples}. \emph{Genome Biol}, \textbf{23}, 79.

\bibitem[\citeproctext]{ref-McGovern:2023}
2. McGovern,K.C., Nixon,M.P. and Silverman,J.D. (2023)
\href{https://doi.org/10.1371/journal.pcbi.1011659}{Addressing erroneous
scale assumptions in microbe and gene set enrichment analysis}.
\emph{PLoS Comput Biol}, \textbf{19}, e1011659.

\bibitem[\citeproctext]{ref-Nixon2024B}
3. Nixon,M.P., Gloor,G.B. and Silverman,J.D. (2024) Beyond
normalization: Incorporating scale uncertainty in microbiome and gene
expression analysis. \emph{bioRxiv},
\href{https://doi.org/10.1101/2024.04.01.587602}{10.1101/2024.04.01.587602}.

\bibitem[\citeproctext]{ref-Dos-Santos:2024aa}
4. Dos Santos,S.J., Copeland,C., Macklaim,J.M., Reid,G. and Gloor,G.B.
(2024) \href{https://doi.org/10.1186/s40168-024-01992-w}{Vaginal
metatranscriptome meta-analysis reveals functional {BV} subgroups and
novel colonisation strategies}. \emph{Microbiome}, \textbf{12}, 271.

\bibitem[\citeproctext]{ref-nixon2024scale}
5. Nixon,M.P., McGovern,K.C., Letourneau,J., David,L.A., Lazar,N.A.,
Mukherjee,S. and Silverman,J.D. (2024)
\href{https://arxiv.org/abs/2201.03616}{Scale reliant inference}.

\bibitem[\citeproctext]{ref-Wu2021}
6. Wu,J.R., Macklaim,J.M., Genge,B.L. and Gloor,G.B. (2021)
\href{https://doi.org/10.1007/978-3-030-71175-7_17}{Finding the centre:
Compositional asymmetry in high-throughput sequencing datasets}. In
Filzmoser,P., Hron,K., Martìn-Fernàndez,J.A., Palarea-Albaladejo,J.
(eds), \emph{Advances in compositional data analysis: Festschrift in
honour of vera pawlowsky-glahn}. Springer International Publishing,
Cham, pp. 329--346.

\bibitem[\citeproctext]{ref-Zhang:2021ab}
7. Zhang,Y., Thompson,K.N., Huttenhower,C. and Franzosa,E.A. (2021)
\href{https://doi.org/10.1093/bioinformatics/btab327}{Statistical
approaches for differential expression analysis in metatranscriptomics}.
\emph{Bioinformatics}, \textbf{37}, i34--i41.

### 2025-04-15 go3-author.tex
cleaned up version

\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (size) of the biological system. Typically a statistical
normalization removes unwanted technical variation in the data or the
parameters of the model to enable differential abundance analyses. We
recently showed that all normalizations make implicit assumptions about
the unmeasured system scale and that errors in these assumptions can
dramatically increase false positive and false negative rates. We
demonstrated that these errors can be mitigated by accounting for
uncertainty using a \emph{scale model}, which we integrated into the
ALDEx2 R package. This article provides new insights focusing on the
application to transcriptomic analysis. We provide transcriptomic case
studies demonstrating how scale models, rather than traditional
normalizations, can reduce false positive and false negative rates in
practice while enhancing the transparency and reproducibility of
analyses. These scale models replace the need for dual cutoff approaches
often used to address the disconnect between practical and statistical
significance. We demonstrate the utility of scale models built based on
known housekeeping genes in complex metatranscriptomic datasets. Thus
this work provides guidance on how to incorporate scale into
transcriptomic data sets.
\end{abstract}

@@

\section{Implementation}\label{implementation}

@@

\singlespacing

>>>>

$^1$corresponding author: ggloor@uwo.ca

@@

\section{Methods}\label{methods}

@@

Acknowledgements: JDS and MPN were supported in part by NIH
1R01GM148972-01. Some aspects of this was supported by an NSERC grant to
GBG.

Availability of Code: All code is available at
\url{https://github.com/ggloor/scale-sim-bio}

Author Contributions: Conceptualization, Methodology, Investigation,
Formal Analysis, Software GBG, MPN, JDS; Visualization GBG, MPN; Writing
Original Draft GBG; Writing Review and Editing GBG, MPN, JDS;
Supervision, Funding JDS. \singlespacing

### 2025-04-15 go3.Rmd
cleaned up version

---
title: "Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor 
    affiliation: Department of Biochemistry, University of Western Ontario 
    email: ggloor@uwo.ca 
  - name: Michelle Pistner Nixon 
    affiliation: Department of Population Health Sciences, Geisinger, Danville, PA 
  - name: Justin Silverman 
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania State University, Department of Medicine, Pennsylvania State University, Institute for Computational and Data Science, Pennsylvania State University, Department of Statistics, Pennsylvania State University 
cite-method: citeproc 
citeproc: TRUE 
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib 
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
  - \usepackage[left]{lineno}

@@

abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system. Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses. We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates. We demonstrated that these errors can be mitigated by accounting for uncertainty using a *scale model*, which we integrated into the ALDEx2 R package. This article provides new insights focusing on the application to transcriptomic analysis. We provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses. These scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance. We demonstrate the utility of scale models built based on known housekeeping genes in complex metatranscriptomic datasets. Thus this work provides guidance on how to incorporate scale into transcriptomic data sets.

@@

# Implementation

@@

\singlespacing

>>>>

---
title: "Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor 
    affiliation: Department of Biochemistry, University of Western Ontario 
    email: ggloor@uwo.ca 
  - name: Michelle Pistner Nixon 
    affiliation: Department of Population Health Sciences, Geisinger, Danville, PA 
  - name: Justin Silverman 
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania State University, Department of Medicine, Pennsylvania State University, Institute for Computational and Data Science, Pennsylvania State University, Department of Statistics, Pennsylvania State University

cite-method: citeproc 
citeproc: TRUE 
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib 
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl 
output:
    pdf_document:
      keep_tex: true
header-includes:
  - \usepackage{setspace}
  - \usepackage[left]{lineno}

@@

Abstract: In high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (size) of the biological system. Typically a statistical normalization removes unwanted technical variation in the data or the parameters of the model to enable differential abundance analyses. We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates. We demonstrated that these errors can be mitigated by accounting for uncertainty using a *scale model*, which we integrated into the ALDEx2 R package. This article provides new insights focusing on the application to transcriptomic analysis. We provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses. These scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance. We demonstrate the utility of scale models built based on known housekeeping genes in complex metatranscriptomic datasets. Thus this work provides guidance on how to incorporate scale into transcriptomic data sets.

@@

# Methods

@@

Acknowledgements: JDS and MPN were supported in part by NIH 1R01GM148972-01. Some aspects of this was supported by an NSERC grant to GBG.

Availability of Code: All code is available at https://github.com/ggloor/scale-sim-bio

Author Contributions: Conceptualization, Methodology, Investigation, Formal Analysis, Software GBG, MPN, JDS; Visualization GBG, MPN; Writing Original Draft GBG; Writing Review and Editing GBG, MPN, JDS; Supervision, Funding JDS.
\singlespacing

### 2025-04-15 go3.tex
cleaned up version

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (size) of the biological system. Typically a statistical
normalization removes unwanted technical variation in the data or the
parameters of the model to enable differential abundance analyses. We
recently showed that all normalizations make implicit assumptions about
the unmeasured system scale and that errors in these assumptions can
dramatically increase false positive and false negative rates. We
demonstrated that these errors can be mitigated by accounting for
uncertainty using a \emph{scale model}, which we integrated into the
ALDEx2 R package. This article provides new insights focusing on the
application to transcriptomic analysis. We provide transcriptomic case
studies demonstrating how scale models, rather than traditional
normalizations, can reduce false positive and false negative rates in
practice while enhancing the transparency and reproducibility of
analyses. These scale models replace the need for dual cutoff approaches
often used to address the disconnect between practical and statistical
significance. We demonstrate the utility of scale models built based on
known housekeeping genes in complex metatranscriptomic datasets. Thus
this work provides guidance on how to incorporate scale into
transcriptomic data sets.
\end{abstract}

>>>>

\begin{document}
\maketitle

### 2025-04-15 supplement_figs.Rmd
cleaned up version

---
title: "Supplement: Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2
"
shorttitle: "Scale ALDEx2 Supp"
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
output:
  pdf_document
---

>>>>

---
title: "Supplement: Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2
"
shorttitle: "Scale ALDEx2 Supp"
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
output:
  pdf_document
header-includes:
  \renewcommand{\figurename}{Supplementary Figure}
---
